Final paper: Towards Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-05-02 paper/paper.tex
paper: draft results part A

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[inline]{enumitem}
\newlist{inlist}{enumerate*}{1}
\setlist[inlist]{label= (\alph*)}
\usepackage[colorlinks]{hyperref}
\hypersetup{%
    colorlinks=true,
    linkcolor=blue,
    urlcolor=cyan,
    citecolor=blue
}
\usepackage[style=ieee]{biblatex}
\addbibresource{sources.bib}

\newcommand{\matlab}{MATLAB}
\newcommand{\mmatlab}{\textmu\matlab}
\newcommand{\colorlib}{CoLoR}
\renewcommand{\algorithmiccomment}[1]{\% #1}
\newcommand{\inputcoq}[1]{\InputIfFileExists{#1.v_tex}{}{\typeout{No file #1.v_tex.}}}
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\func}[1]{\mathsf{#1}}

@@

>>>>

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{array,booktabs}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[inline]{enumitem}
\newlist{inlist}{enumerate*}{1}
\setlist[inlist]{label= (\alph*)}
\usepackage{mathpartir,semantic}
\usepackage[colorlinks]{hyperref}
\hypersetup{%
    colorlinks=true,
    linkcolor=blue,
    urlcolor=cyan,
    citecolor=blue
}
\usepackage[style=ieee]{biblatex}
\addbibresource{sources.bib}

\newcommand{\matlab}{MATLAB}
\newcommand{\mmatlab}{\textmu\matlab}
\newcommand{\colorlib}{CoLoR}
\renewcommand{\algorithmiccomment}[1]{\% #1}
\newcommand{\inputcoq}[1]{\InputIfFileExists{#1.v_tex}{}{\typeout{No file #1.v_tex.}}}
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\func}[1]{\mathsf{#1}}
\newcommand{\iname}[1]{\textsf{#1}}

@@

\section{Results}\label{S:results}

In this section, we present the results of our work towards proving termination
of our mosaic algorithm. As noted, our primary accomplishment is the definition
of a workable matrix type and operations on it. This matrix type is intended to
support the operations needed to implement expression evaluation for \mmatlab, a
crucial component of implementing the language's semantics. It remains to finish
implementing expression evaluation and the necessary matrix operations, to give
operation semantics for the statements of \mmatlab, to give a total Hoare logic
for the language, and then finally to give the proofs of termination of the
algorithms found in \figurename{s}~\ref{alg:v1},~\ref{alg:v2}, and~\ref{alg:v3}.

We present, in order,
\begin{inlist}
\item the Coq definition of the polymorphic matrix type \texttt{matrix A} and
    corresponding operations;
\item key theorems and sketches of their proofs about the matrix type and its
    operations; and
\item advanced tactics, automation, and ideas for Coq proofs used in the
    development.
\end{inlist}

% TODO: update zelda_mosaic_proof version with final commit

All code may be found at the project repository~\cite{zelda_mosaic_proof}. Most
of the work in this section corresponds to the file \texttt{Lib Matrix.v}.

\subsection{Defining Matrices}\label{S:matrix_defn}

First, we define the matrix type. After noting that it allows invalid matrices
to be formed, we define the well-formed property on matrices in two equivalent
ways. Finally, we give an overview of the Coq definitions of three operations on
matrices: computing their shape, linearizing, and linear indexing. We briefly
mention the beginning of our work on multi-dimensional and range-based indexing.

\subsubsection{The matrix type}

As mentioned in \S~\ref{S:design_dependent}, our original formulation of
matrices as nested dependent \texttt{Vector.t}s quickly grew in complexity.
We abandoned that approach in favor of a type that is not guaranteed to behave
well, but whose definition makes defining operations more
natural~\cite{SO_2021_2}. This approach removes the proofs that dependent types
provide automatically, so we have to give those ourselves.

The new design pairs the dimensionality of a matrix, called its \emph{shape},
with its \emph{contents}. The pairing is done via \emph{Record} types (see the
Coq manual~\cite{Coq}), which are tuples with named elements. The definition
follows below:

\begin{align*}
    & \func{matrix} (\var{A}: \var{Type}) \triangleq \\
    & \{
        \func{shape}: \func{list}\, \mathbb{N},
        \func{contents}: \func{matrix\_content}\, \var{A}
    \}
\end{align*}

The shape is a list of naturals. Each natural corresponds (in a well-formed
matrix) to the number of elements of the matrix in that dimension. For example,
a vector of five (5) elements has shape \texttt{[5]}, while a two-dimensional
matrix of three-by-two elements has shape \texttt{[3; 2]}. The vacuous case,
when the shape is the empty list, is a single scalar value: there is no matrix.

In parallel, the contents are wrapped in an inductive polymorphic type
\(\func{matrix\_content}\, \var{A}\). The type's two constructors correspond to
the two cases above: a \(\func{Matrix}\) contains a list of
\(\func{matrix\_content}\, \var{A}\) values. The \(\func{Matrix}\) constructor,
in a well-formed matrix, corresponds to one dimension of the shape. A
\(\func{Scalar}\) contains exactly a value of type \(\var{A}\) and is the
conceptually the base-case for the matrix type.

\begin{align*}
    & \func{matrix\_content} (\var{A}: \var{Type}) \triangleq\; \\
    & |\; \func{Scalar}: \var{A} \to \func{matrix\_content}\, \var{A} \\
    & |\; \func{Matrix}: \func{list}\, (\func{matrix\_content}\, \var{A}) \to \func{matrix\_content}\, \var{A}
\end{align*}

We provide a few concrete examples of mathematical objects in the space of
scalars and multi-dimensional matrices containing a single type of object in
Table~\ref{T:matrices}. Each example shows standard mathematical notation, the
Coq notation using our matrix type, and the resulting type of the value in Coq.

\begin{table*}[ht]
    \centering
    \begin{tabular}{c >{\ttfamily}c >{\ttfamily}c}
        Mathematical notation & Coq notation & Coq type \\ \toprule
        \(5\) & \{| shape := []; contents := Scalar 5 |\} & matrix nat \\
        \(\begin{bmatrix}
            1 & 2 & 3
        \end{bmatrix}\)
        & \{| shape := [3]; contents := Matrix [Scalar 1; Scalar 2; Scalar 3] |\} & matrix nat \\
        \\
        \(\begin{bmatrix}
            1 & 2 \\
            3 & 4 \\
            5 & 6
        \end{bmatrix}\)
        & \{| shape := [3; 2]; contents := Matrix [
        & matrix nat \\
        & Matrix [Scalar 1; Scalar 2]; & \\
        & Matrix [Scalar 3; Scalar 4]; & \\
        & Matrix [Scalar 5; Scalar 6]] |\} &
    \end{tabular}
    \caption{Example matrices and their Coq equivalents}\label{T:matrices}
\end{table*}

\subsubsection{The well-formed property}

As suggested above, the lack of dependent typing means a value of type
\(\func{matrix}\, \var{A}\) is \emph{not} guaranteed to be well-formed. All of
the following are valid but nonsensical values:

\begin{verbatim}
{| shape := [];
   contents := Matrix [Scalar 1] |}
{| shape := [2];
   contents = Matrix [Scalar 2] |}
{| shape := [1; 2; 3; 0];
   contents = Scalar 0 |}
\end{verbatim}

In the first case, we have a matrix with no dimensions, which should be a scalar
and is not. In the second case, we have a one-dimensional matrix of two
elements which only contains one element. In the third, we have an empty
four-dimensional matrix which is actually a scalar. None of these pairings makes
sense---we need a way to exclude them from our results, effectively categorizing
these matrices as ``not well-formed.'' Conversely, we would like a way to
categorize matrices as well-formed.

We have informally given pieces of the definition in laying our the relationship
between shape and contents; here, we make the definition more precise. A matrix
is well-formed if, and only if,
\begin{itemize}
    \item the shape is an empty list, and the contents are a single scalar value;
        or
    \item the shape is a non-empty list consisting of a head \(n\) representing
        the length of the first dimension and tail \(t\) representing the shape
        of the subsequent dimensions, and the contents is a \(\func{Matrix}\)
        containing a list with length \(n\), and each element of the list is
        well-formed with respect to \(t\).
\end{itemize}
In Coq, we express this definition via a recursive definition
\(\func{well\_formed'}\) which computes a proposition, or proof-burden, over a
shape and contents. We wrap this up in a definition \(\func{well\_formed}\),
which unpacks a matrix for use with \(\func{well\_formed'}\). With this
definition, we can theorems in terms of matrices that are assumed to be
well-formed or that must be shown to be well-formed.

In the course of our development, we discovered that this computational version
did not always provide a convenient means to work a proof. Clearly, the
recursive nature of matrices and well-formedness means that most proofs will be
by induction. But induction on what? There is no induction principle for the
record type, so we must choose from induction on the shape or on the contents.
It turns out that there is a natural inductive version of the well-formed
property that permits induction on evidence. In \S~\ref{S:matrix_thm} we will
state and sketch the equivalence theorem between the two definitions. For now,
we give the judgment rules for \(\func{well\_formedI'}\), the inductive
counterpart to \(\func{well\_formed'}\) in \figurename~\ref{F:wfI}. We wrap the
inductive proposition in \(\func{well\_formedI}\), which unpacks a matrix the
same way as \(\func{well\_formed}\).

\begin{figure*}[ht]
    \centering
    \begin{mathpar}
        \inference[\iname{WF\_Scalar}]{%
            a: A
        }{\func{well\_formedI'}\, []\, (\func{Scalar}\, a)}

        \and

        \inference[\iname{WF\_Empty}]{%
            t: \func{list}\, \mathbb{N}
        }{\func{well\_formedI'}\, (0::t)\, (\func{Matrix}\, [])}

        \and

        \inference[\iname{WF\_Cons}]{%
            h: \mathbb{N} &
            t: \func{list}\, \mathbb{N} &
            m: \func{matrix\_contents}\, \var{A} \\
            ms: \func{list}\, (\func{matrix\_content}\, \var{A}) &
            \func{well\_formedI'}\, (h::t)\, (\func{Matrix}\, ms) &
            \func{well\_formedI'}\, t\, m
        }{\func{well\_formedI'}\, (1+h::t)\, (\func{Matrix}\, (m::ms))}

    \end{mathpar}
    \caption{Inductive judgments for the well-formed property of
    matrices}\label{F:wfI}
\end{figure*}

The important features of the judgements are \(\iname{WF\_Emtpy}\) and
\(\iname{WF\_Cons}\). The judgement \(\iname{WF\_Empty}\) corresponds to the
vacuous case: all elements in an empty list are well-formed with respect to any
shape (there are no such elements!). The judgment \(\iname{WF\_Cons}\) provides
a judgment for increasing the size of the first dimension of a well-formed
matrix by adding a well-formed component to the contents.

It is easy to prove that all the matrices in Table~\ref{T:matrices} are
well-formed with respect to both definitions. It is likewise easy to prove that
the previously-sketched invalid examples are not well-formed under both
definitions.

\subsubsection{Some operations on matrices}

The operations we have defined are computing the shape of a matrix, linearizing
the matrix, and indexing a multi-dimensional matrix linearly. We will discuss
these and give a brief note about our approach to multi-dimensional range-based
indexing.

\paragraph{Computing the shape of a matrix} While each matrix carries its shape
as part of the record pairing, it is occasionally convenient to compute from the
contents alone the shape. There are two subtleties worth mentioning: first, the
computation is only valid for well-formed matrices because it relies on
assumptions about the relationship between different components of the contents.
Second, the computed shape may not be the most general: we have already seen
that, when one dimension in a matrix has shape zero, the subsequent dimensions
can be arbitrarily sized and the result is still well-formed! This impacts the
correctness statement for shape-computation: on a well-formed matrix, it may not
give the exact same shape, yet it will give a shape such that the original
matrix is well-formed with respect to the new shape. We can call this
``normalization'' of the shape, as it discards dimensions that follow a
zero-sized dimension. See \S~\ref{S:matrix_thm} for more details on
normalization via shape computation. The definition is straightforward: the
function \(\func{compute\_shape}\) yields
\begin{itemize}
    \item an empty list when the contents consist of a single scalar;
    \item a singleton list containing zero when the contents consist of an empty
        matrix; and
    \item a list with head the length of the matrix contents and tail the
        computed shape of just one of the contents when the contents
        consist of a non-empty matrix.
\end{itemize}
This last bullet explains why the function is only valid on well-formed
matrices: it implicitly assumes that all the elements of the matrix contents
have the same computed shape, which is only true when the matrix is well-formed.
The same definition is given below:

\begin{align*}
    & \func{compute\_shape}(\var{m}) \triangleq \\
    & \begin{cases}
        [] & m = \func{Scalar}\, a \\
        [0] & m = \func{Matrix}\, [] \\
        |\var{ms}| :: \func{compute\_shape}\, m' & m = \func{Matrix}\, \var{ms}, \\
        & \var{ms} = m'::t
    \end{cases}
\end{align*}

\paragraph{Linearizing a matrix} See \S~\ref{S:design_linearize} for more
details on matrix linearization. To define linearization in Coq, we use a
two-stage unpacking as with other matrix operations. The function
\(\func{linearize'}\) turns scalars into a singleton list and recursively
flat-maps itself over matrices to produce the one-dimensional matrix as a list.
The function \(\func{linearize}\) unpacks a matrix, then transforms the contents
by wrapping the result of mapping \(\func{linearize'}\) into \(\func{Scalar}\)s
in a \(\func{Matrix}\) and transforms the shape by taking the product of the
(normalized) shape of the matrix. Recall that the length of the linearized
vector is the product of the sizes of all the dimensions. The definitions are
spelled out below:

\begin{align*}
    & \func{linearize'}(\var{contents}) \triangleq \\
    & \begin{cases}
        [\var{a}] & \var{contents} = \func{Scalar}\, \var{a} \\
        \func{flat\_map}\, \func{linearize'}\, \var{ms} & \var{contents} =
        \func{Matrix}\, ms
    \end{cases} \\
    & \func{linearize}(\{\var{contents}, \var{shape}\}) \triangleq \\
    & \{ \var{shape} = \left[\prod \func{compute\_shape}\, \var{contents}\right]; \\
    & \ \, \var{contents} = \func{Matrix}\, (\func{map}\, \func{Scalar}\, (\func{linearize'}\, \var{contents})) \}
\end{align*}

\paragraph{Linearly indexing a matrix} Given the operation to linearize a
matrix, one-based linearly indexing is easy. See \S~\ref{S:design_linear_index}
for details on linear indexing. The function \(\func{nth}\) takes a matrix \(m\)
and an index \(n\); it returns the one-based linearly indexed content of \(m\)
at \(n\) (wrapped in the option constructor \(\func{Some}\)) if \(n\) is in
bounds relative to \(m\). When \(n\) is out of bounds, the result is
\(\func{None}\). For one-based indexing with functions that naturally use
zero-based indexing, we use \(n-1\). We should note that the current version of
\(\func{nth}\) has the subtle bug that indexing at \(n=0\) currently returns the
same as indexing at \(n=1\), when it should in fact be out of bounds and thus
\(\func{None}\)---a simple \texttt{if} or \texttt{match} guard would cover this
case.

\subsection{Key Theorems about Matrices}\label{S:matrix_thm}

\subsection{Advanced Coq for Matrices}\label{S:matrix_coq}
